Analysis and elimination of bottlenecks in parallel algorithm for solving global optimization problems

### 2022-04-13 2022/RSD/paper.tex
fix some formulaе and references

\usepackage{hyperref}
%
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
\usepackage{color}
\renewcommand\UrlFont{\color{blue}\rmfamily}
\usepackage{amsmath} %?????????!!!!!!!
%
\begin{document}
%
\title{Analysis and elimination of bottlenecks in parallel algorithm for solving global optimization problems}
%
\titlerunning{Analysis and elimination of bottlenecks in parallel algorithm}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{
Konstantin Barkalov \orcidID{0000-0001-5273-2471} \and
Ilya Lebedev \orcidID{0000-0002-8736-0652} \and
Denis Karchkov \orcidID{0000-0002-4164-5927}
}
%
\authorrunning{K. Barkalov, I. Lebedev, D. Karchkov}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Lobachevsky State University of Nizhny Novgorod, Nizhny Novgorod, Russia \\
\email{\{konstantin.barkalov,ilya.lebedev,karchkov\}@itmm.unn.ru}
}

@@

Let us consider a problem of searching the global minimum of an $N$-dimensional function $\varphi(y)$ in a hyperinterval $D$
\begin{equation}\label{problemN}
\varphi(y^*) = \min{\{ \varphi(y):y \in D \}}
\end{equation}
\begin{equation}\label{D}
D = \{ y \in R^{N}: a_i \leq y_i \leq b_i, 1 \leq i \leq N \}.
\end{equation}

The main assumption, which the algorithm considered are based on is the fulfillment of Lipschitz condition for the function $\varphi(y)$, i.e.
$$ \mid \varphi(y_1) - \varphi(y_2) \mid \leq L\|y_1 - y_2\|, y_1, y_2 \in D, 0 < L < \infty. \eqno(3)$$
Note that the Lipschitz constant $L$ featuring the rate of variation of the function is not known \textit{a priori}. This assumption is true for a wide class of problems and is typical of many optimization techniques (see, e.g., \cite{Evtushenko2013,Jones2009,Paulavicius2014}).
%of finding the optimal values of the parameters of mathematical models from the experimental data.

The known approaches to solving the Lipschitz global optimization problems are based on a generalization of the efficient algorithms for solving the one-dimensional problems onto solving the multidimensional ones (see, for example, the methods of diagonal partitions \cite{Sergeyev2017} or simplicial partitions \cite{PaulaviciusZilinskas2014}). In the present work, we have also used the approach reducing the solving of a multidimensional problem to solving corresponding one-dimensional one. This approach is based on the idea of the dimensionality reduction using Peano-Hilbert curve $y(x)$ mapping the interval $[0,1]$ of the real axis onto an $N$-dimensional cube 
$$ \{ y \in R^{N}: -2^{-1} \leq y_i \leq 2^{-1}, 1 \leq i \leq N\} = \{y(x): 0 \leq x \leq 1 \}.    \eqno(4)$$

@@

The use of the space-filling curves allows reducing the multidimensional problem \ref{problemN} to a one-dimensional problem 
$$\varphi(y^*) = \varphi(y(x^*)) = \min{\{ \varphi(y(x)): x \in [0, 1] \}}. \eqno(5)$$
and applying the efficient one-dimensional optimization methods to solve this one. Keeping the property of limited relative differences of the objective function is an important property of using the mappings of this kind. If the function $\varphi(y)$ in the domain $D$ satisfies Lipschitz condition, the function $\varphi(y(x))$ in the interval $[0,1]$ will satisfy uniform H{\"o}lder condition
$$\mid \varphi(y(x_1)) - \varphi(y(x_2)) \mid \leq H {\mid x_1 - x_2 \mid}^{1/N}, x_1, x_2 \in [0,1], \eqno(6)$$
where H{\"o}lder constant $H$ is related to Lipschitz constant $L$ by the relation $H = 2L\sqrt{N+3}$. Therefore, without liming the generality, one can consider the minimization of the one-dimensional function 
$$ f(x) = \varphi(y(x)), x \in [0,1], \eqno(7)$$
satisfying H{\"o}lder condition.

@@

Step 1. Renumber the points of the previous trials $\{x^0,...,x^k\}$ by the lower indices in the order of increasing the coordinate values, i.e. 
$$ 0 = x_0 < x_1 < ... < x_k = 1.$$
Step 2. Calculate the magnitudes 
$$ \mu = \max\limits_{1 \leq i \leq k} \frac{\mid z_i - z_{i-1} \mid}{\Delta_i}, $$

\begin{displaymath}
M = \left\{ \begin{array}{ll}
                r\mu, & \textrm{$\mu > 0$,}\\
                1, & \textrm{$\mu = 0$}.
  \end{array} \right.
\end{displaymath}
where $z_i = f(y(x_i)),$  $\Delta_i = {(x_i - x_{i-1})}^{1/N}$, and $r > 1$ is a predefined parameter of the method.

Step 3. For each interval $(x_{i-1}, x_i)$, $1 \leq i \leq k$ compute the characteristic according to the formula
$$ R(i) = \Delta_i + \frac{(z_i - z_{i-1})^2}{M^2 \Delta_i} - 2 \frac{z_i + z_{i-1}}{M}, 1 \leq i \leq k. $$

Step 4. Arrange the characteristics $R(i)$, $1 \leq i \leq k$ in the decreasing order
$$ R(t_1) \geq R(t_2) \geq ... \geq R(t_{k-1}) \geq R(t_{k+1}) $$
and select $P$ largest characteristics with the interval indices $t_j$, $1 \leq j \leq P$.

Step 5. Perform new trials at the points $x^{k+j}$, $1 \leq j \leq P$ computed by the formulae
$$ x^{k+1} = \frac{x_{t_j} + x_{t_j - 1}}{2} - sign(z_{t_j} - z_{t_j - 1}) \frac{1}{2r} {\Bigg[\frac{\mid z_{t_j} - z_{t_j - 1} \mid}{\mu}\Bigg]}^{N}, 1 < t_j < k+1.$$
Put the results of the trials performed into the information database of the algo-rithm.\\

Step 6. Check the stopping criterion $\Delta_{t_j} \leq \varepsilon$ where $t_j$ from \cite{} and $\varepsilon > 0$ is the predefined precision.

After completing the algorithm work, take as the estimate of the solution of the problem \ref{problemN} 
$$ f_k^* = \min\limits_{1 \leq i \leq k} f(x^i), x_k^* = \arg \min\limits_{1 \leq i \leq k} f(x^i).$$

The theoretical substantiation of the algorithm presented here see in more details in \cite{}. Here, note briefly that the characteristics of the intervals used in the algorithm \cite{} can be considered as a quantitative measure how particular interval is promising from the viewpoint of finding the global minimum inside this one. The inequalities \cite{} arrange the intervals according to the characteristics on the ones, and the trials are performed in parallel in p most promising intervals.

\subsubsection{Theoretical estimates of the speedup of the parallel algorithm}
Let us describe (according to \cite{}) the theoretical properties of the parallel algorithm featuring its speedup. In the problems being solved, the time of executing a single search trial exceeds the time of processing its results essentially. Therefore, the speedup in the number of trials is the key characteristic of the efficiency of the parallel global search algorithm
$$s(p) = k(1)\frac{p}{k(p)}, \eqno(26)$$
where $k(1)$ is the number of trials executed by the sequential method and $k(p)$ is the number of trials executed by the parallel method employing p parallel processes.\\

Obviously, the quantities of trials $n(p)$ for the algorithms with different degrees of parallelization $p$ will differ from each other. Indeed, the sequential algorithm has the full information obtained at the previous $k$ iterations when selecting the point $x^{k+1}$ of the next $(k+1)^{th}$ trial. The parallel algorithm selects not one but $p$ points $x^{k+j}$, $1 \leq j \leq p$ on the base of the same information at the $(k+1)^{th}$ iteration. It means that the selection of the points $x^{k+j}$ is performed without the information on the trial results at the points $x^{k+i}$, $1 \leq i < j$. Only the first point $x^{k+1}$ will match to the point selected by the sequential algorithm. The points of other trials, in general, may not match the ones generated by the sequential algorithm. Therefore, we will call such trials redundant and the quantity
\begin{displaymath}
\lambda(p) = \left\{ \begin{array}{ll}
                (n(p) - n(1)) / n(p), & \textrm{$n(p) > n(1)$}\\
                0, & \textrm{$n(p) \leq n(1)$}
  \end{array} \right.
\end{displaymath}
we will call the redundancy of the method.

@@

\textbf{Theorem}. Let $x^*$ be the global minimum point and $x^{\prime}$ be the local minimum one of the function $f(x)$ and let the following conditions be satisfied:
    \begin{enumerate}
        \item The inequality is fulfilled
        $$ f(x^{\prime}) - f(x^*) \leq \delta, \delta > 0. \eqno(28)$$
        \item The first $q(l)$ trials of the sequential method and of the parallel one match to each other i. e.
        $$ \{x^1,...,x^{q(l)}\} = \{y^1,...,y^{q(l)}\} \eqno(29)$$
        where
         $$ \{x^1,...,x^{q(l)}\} \subset \{x^k\}, \{y^1,...,y^{q(l)}\}\subset \{y^k\}. \eqno(30)$$
        \item There exist a point $t^n \in \{y^m\}$, $n < q(l)$ such as $x^{\prime} \leq y^n \leq x^*$ or $x^* \leq y^n \leq x^{\prime}$.
        \item For the quantity $M$ from \cite{}, the inequality is fulfilled 
        $$ M > 2^{2 - 1/N} H. \eqno(31)$$
        where $H$ is Hölder constant of the minimized one-dimensional function.
    \end{enumerate}
Then, the parallel algorithm at $p=2$ will be non-redundant (i. e. $s(2)=2$, $\lambda(2)=0$) as long as the condition is fulfilled 
$$ (x_{t_j} - x_{t_j - 1})^{1/N} > \frac{4\delta}{M - 2^{2 - 1/N} H}, j=1,2. \eqno(32)$$
where $t_j$ is defined according to \cite{}.\\
\textbf{Consequence}. Let the objective function $f(x)$ have $Q$ points of local minima $\{x_1^{\prime},...,x_Q^{\prime}\}$, which the condition (28) is fulfilled for, and let there exist the trial points $y^{n_i}$, $1 \leq i \leq Q$ such as 
$$ y^{n_i} \in \{y^1,...y^{q(l)}\}, \eqno(33)$$
$$ \alpha_i \leq y^{n_i} \leq \alpha_{i+1}, \alpha_i, \alpha_{i+1} \in \{x^*, x_1^{\prime},...,x_Q^{\prime}\}, 1 \leq i \leq Q. \eqno(16)$$
then, if the conditions of Theorem are fulfilled, the parallel algorithm with the parallelization degree $Q+1$ will be non-redundant (i. e. $s(Q+1)=Q+1, \lambda(Q+1)=0$) until the condition (32) is fulfilled.\\
The consequence from Theorem plays a special role in solving the multidimen-sional problems reduced to the one-dimensional ones using the Peano-like evolvent $y(x)$. The evolvent $y(x)$, being an approximation to the Peano curve, has the effect of «splitting» of the global minimum point $y^* \in D$ into several (up to $2^N$) preimages in the interval [0,1]. Applying the parallel algorithm to minimize the reduced one-dimensional function, one can expect a zero redundancy at the parallelization degree up to $2^N+1$.\\
Now, let us consider the case when after the stage of the search for the rough approximation to the global solution, the stage of its local refinement is performed. Assume that the fraction of the search trials necessary for the local refinement of the solution makes $\alpha$ of total number of trials. Then, according to Amdahl's law, the overall speedup $S(p)$, which can be obtained by parallelization into $p$ processes, cannot exceed the quantity 
$$ S(p) = {\Big(\alpha + \frac{1-\alpha}{p}\Big)}^{-1}.$$
This relation will be true if one assumes the non-redundant parallelization of the global search stage. The increase in the efficiency of the parallel algorithm will be limited by the fraction of trials at the stage of the local refinement of the solution. So far, to increase the efficiency of parallelization (especially in the case of a large number of processes employed), one needs to parallelize not only the global search but the local one as well. Two particular local methods applicable to the optimization problems with the «black box» type functions as well as the methods of parallelization of these ones will be considered in the next section.

\section{Local optimization methods}
\subsection{BFGS method}

BFGS (Broyden-Fletcher-Goldfarb-Shanno) method (see [Nocedal, Wright]) belongs to the class of quasi-Newton methods using the values of the function $f(x^k)$ and of its gradient $\nabla f(x^k)$ to organize the search. In this method, the function f is assumed to have the properties of the quadratic model that corresponds to the existence of a symmetric matrix of the second derivatives (Hessian) of the function $\nabla^2 f(x^k)$.\\
The basic idea for all quasi-Newton methods consists in the construction of an ap-proximation of the actual Hessian $\nabla^2 f(x^k)$ at every step of the algorithm using a special matrix $B^k$ followed by selection of the descent direction $d^k$ as a solution of the following system:
$$ B^k d^k = -\nabla f(x^k) \leftrightarrow d^k = -H^k \nabla f(x^k),$$
where $H^k={(B^k)}^{-1}$.\\
After obtaining the direction, a step is performed from current point $x^k$:
$$ x^{k+1}=x^k+ \alpha_k d^k,$$
where $\alpha_k>0$ is the adjustable step length.\\
Assuming $B^0=E$, the Hessian approximation is recalculated by the formula $B^{k+1}= B^k+ U^k$ where $U^k$ are the corrections represented in the form of a matrix of the rank 1. The small rank of the corrections $U^k$ is necessary for efficient procedure of calculating the inverse matrix $H^{k+1}=(B^{k+1})^{-1}=(B^k+ U^k)^{-1}$.\\
The main requirement to forming the corrections $U^k$ of particular form is the secant equation $B^{k+1} (x^{k+1} - x^k )= \nabla f(x^{k+1})- \nabla f(x^k)$, which is performed for all quasi-Newton methods. For more precise determination of the matrix $B^{k+1}$, it is necessary to impose additional requirements on the Hessian approximation. The set of such requirements is determined by BFGS rule. Assume:
$$ s_k = x^{k+1} - x^k,$$
$$ y_k = \nabla f(x^{k+1})- \nabla f(x^k).$$
Then, the BFGS recalculation scheme takes the following form:
$$ B^{k+1} = B^k - \frac{B^k s_k s_k^T B^k}{s_k^T B^k s_k} + \frac{y^k (y^k)^{T}}{(y^k)^{T} s_k},$$
$$ H^{k+1} = (E - \rho_k s_k (y^k)^{T}) H^k (E - \rho_k y^k s_k^T) + \rho_k s_k s_k^T,$$
where $\rho_k = 1/((y^k)^{T} s_k)$.\\
For the systems with limited memory, a modification of BFGS method – the lim-ited BFGS in the limited space (L-BFGS-B) is used. The modified method stores the history of $\aleph^k = ((s_{k-i}, y^{k-i}))_i^l$ of the last $l$ vectors $s_k$ and $y^k$. From the history, the approximate form of the matrix $H^k$ is restored.\\
The algorithm of the BFGS method can be represented by a sequence of the fol-lowing commands:

@@

Step 6. Stop if the condition $\|\nabla f^k\| < \varepsilon$ is fulfilled.

\subsubsection{Hooke-Jeeves method}
Hooke-Jeeves method (see, e.g., \cite{Himmelblau} ) is the first-order one, i.e. it needs only the function values for the work. In this method, the search for the minimum at every step is performed as a result of a shift along some sample direction (step by a sample), which is constructed and then corrected as a result of special trial coordinate translations called constructing a configuration.

@@

if $\varphi(\bar{y} + he^i) < \varphi(\bar{y})$ then set $\bar{y} = \bar{y} + he^i$ else if $\varphi(\bar{y} - he^i) < \varphi(y)$ then $\bar{y} = y - he^i.$
Next, the step-by-step description of the whole method is given.

@@

The meaning of the actions performed within the steps 2, 3, and 4 can be explained as follows. \\
Step 4 was introduced in order to enable the method to increase the shift magnitude within a single iteration rapidly in the case when the point $t^k$ is located far away from the solution. At that, prior to make $y^{k+1}$ current point of the iteration, a configuration is constructed from the point $y^{k+1}$. By these means, in the case of obtaining a point with $\varphi(t^{k+1})< \varphi(t^k)$, the next step by sample, in general, will be performed in a changed direction (relative to previous one) that allows adapting the method to variations in the function relief.\\
Finally, upon obtaining a value $\varphi(t^{k+1}) \geq \varphi(t^k)$ at Step 3, an attempt to run the algorithm again from the point $t^0=t^k$ (better than the one found) is undertaken. At that, if such an attempt hasn’t been undertaken yet, the parameter $h$ is not changed, but if it has been undertaken already, $h$ is cut in half first.

\subsubsection{Parallelization of the local optimization methods}

@@

For example, in Hooke-Jeeves method one can parallelize the construction of configuration i. e. parallelize the loop at the second step of the algorithm. In this loop, one may not compute the function values at the points $\bar{y} + he^i$ or $\bar{y} - he^i$ immedeatedly but may store the coordinates of the ones in the buffer, and upon accumulating P points compute the function values at these ones in parallel. Afterwards, the accumulated information is analyzed and the coordinates of the point $\bar{y}$ are changed. This operation is repeated until all $N$ coordinates $\bar{y}$ are computed.

@@

>>>>

\usepackage{hyperref}
%
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
\usepackage{color}
\renewcommand\UrlFont{\color{blue}\rmfamily}
\usepackage{amsmath} 
\usepackage{multirow} 
%
\begin{document}
%
\title{Analysis and elimination of bottlenecks in parallel algorithm for solving global optimization problems}
%
\titlerunning{Analysis and elimination of bottlenecks in parallel algorithm}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{
Konstantin Barkalov \orcidID{0000-0001-5273-2471} \and
Ilya Lebedev \orcidID{0000-0002-8736-0652} \and
Denis Karchkov \orcidID{0000-0002-4164-5927}
}
%
\authorrunning{K. Barkalov, I. Lebedev, D. Karchkov}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Lobachevsky State University of Nizhny Novgorod, Nizhny Novgorod, Russia \\
\email{\{konstantin.barkalov,ilya.lebedev,karchkov\}@itmm.unn.ru}
}

@@

Let us consider a problem of searching the global minimum of an $N$-dimensional function $\varphi(y)$ in a hyperinterval $D$
\begin{gather}
\varphi(y^*) = \min{\{ \varphi(y):y \in D \}} \label{problemN} \\ 
D = \{ y \in R^{N}: a_i \leq y_i \leq b_i, 1 \leq i \leq N \}. \label{D}
\end{gather}

The main assumption, which the algorithm considered are based on is the fulfillment of Lipschitz condition for the function $\varphi(y)$, i.e.
\begin{equation} \label{lip_ref}
\mid \varphi(y_1) - \varphi(y_2) \mid \leq L\|y_1 - y_2\|, y_1, y_2 \in D, 0 < L < \infty.
\end{equation}

Note that the Lipschitz constant $L$ featuring the rate of variation of the function is not known \textit{a priori}. This assumption is true for a wide class of problems and is typical of many optimization techniques (see, e.g., \cite{Evtushenko2013,Jones2009,Paulavicius2014}).
%of finding the optimal values of the parameters of mathematical models from the experimental data.

The known approaches to solving the Lipschitz global optimization problems are based on a generalization of the efficient algorithms for solving the one-dimensional problems onto solving the multidimensional ones (see, for example, the methods of diagonal partitions \cite{Sergeyev2017} or simplicial partitions \cite{PaulaviciusZilinskas2014}). In the present work, we have also used the approach reducing the solving of a multidimensional problem to solving corresponding one-dimensional one. This approach is based on the idea of the dimensionality reduction using Peano-Hilbert curve $y(x)$ mapping the interval $[0,1]$ of the real axis onto an $N$-dimensional cube 
\begin{equation} \label{n_dem_cube_ref}
\{ y \in R^{N}: -2^{-1} \leq y_i \leq 2^{-1}, 1 \leq i \leq N\} = \{y(x): 0 \leq x \leq 1 \}.
\end{equation}

@@

The use of the space-filling curves allows reducing the multidimensional problem \ref{problemN} to a one-dimensional problem 
\begin{equation} \label{problem_ref}
\varphi(y^*) = \varphi(y(x^*)) = \min{\{ \varphi(y(x)): x \in [0, 1] \}}
\end{equation}
and applying the efficient one-dimensional optimization methods to solve this one. Keeping the property of limited relative differences of the objective function is an important property of using the mappings of this kind. If the function $\varphi(y)$ in the domain $D$ satisfies Lipschitz condition, the function $\varphi(y(x))$ in the interval $[0,1]$ will satisfy uniform H{\"o}lder condition
\begin{equation} \label{hölder_ref}
\mid \varphi(y(x_1)) - \varphi(y(x_2)) \mid \leq H {\mid x_1 - x_2 \mid}^{1/N}, x_1, x_2 \in [0,1],
\end{equation}
where H{\"o}lder constant $H$ is related to Lipschitz constant $L$ by the relation $H = 2L\sqrt{N+3}$. Therefore, without liming the generality, one can consider the minimization of the one-dimensional function 
\begin{equation} \label{minim_fun_ref}
f(x) = \varphi(y(x)), x \in [0,1],
\end{equation}
satisfying H{\"o}lder condition.

@@

Step 1. Renumber the points of the previous trials $\{x^0,...,x^k\}$ by the lower indices in the order of increasing the coordinate values, i.e. 
$$ 0 = x_0 < x_1 < ... < x_k = 1.$$
Step 2. Calculate the magnitudes 
\begin{equation}\label{M}
\mu = \max\limits_{1 \leq i \leq k} \frac{\mid z_i - z_{i-1} \mid}{\Delta_i}, \;\;
M = \left\{ \begin{array}{ll}
                r\mu, & \textrm{$\mu > 0$,}\\
                1, & \textrm{$\mu = 0$},
  \end{array} \right.
\end{equation}
where $z_i = f(y(x_i)),$  $\Delta_i = {(x_i - x_{i-1})}^{1/N}$, and $r > 1$ is a predefined parameter of the method.

Step 3. For each interval $(x_{i-1}, x_i)$, $1 \leq i \leq k,$ compute the \textit{characteristic} according to the formula
\begin{equation}\label{R}
R(i) = \Delta_i + \frac{(z_i - z_{i-1})^2}{M^2 \Delta_i} - 2 \frac{z_i + z_{i-1}}{M}, 1 \leq i \leq k. 
\end{equation}

Step 4. Arrange the characteristics $R(i)$, $1 \leq i \leq k,$ in the decreasing order
\begin{equation} \label{OrderedR}
 R(t_1) \geq R(t_2) \geq ... \geq R(t_{k-1}) \geq R(t_{k+1}) 
\end{equation}
and select $P$ largest characteristics with the interval indices $t_j$, $1 \leq j \leq P$.

Step 5. Perform new trials at the points $x^{k+j}$, $1 \leq j \leq P$ computed by the formulae
$$ x^{k+1} = \frac{x_{t_j} + x_{t_j - 1}}{2} - sign(z_{t_j} - z_{t_j - 1}) \frac{1}{2r} {\Bigg[\frac{\mid z_{t_j} - z_{t_j - 1} \mid}{\mu}\Bigg]}^{N}.$$
Put the results of the trials performed into the information database of the algorithm.

Step 6. Check the stopping criterion $\Delta_{t_j} \leq \varepsilon$, where $t_j$ are from Step 4 and $\varepsilon > 0$ is the predefined precision. 
After completing the algorithm work, take as the estimate of the solution of the problem \ref{problemN} 
$$ f_k^* = \min\limits_{1 \leq i \leq k} f(x^i), x_k^* = \arg \min\limits_{1 \leq i \leq k} f(x^i).$$

The theoretical substantiation of the algorithm presented here see in more details in \cite{Strongin2000}. Here, note briefly that the characteristics of the intervals (\ref{R}) used in the algorithm can be considered as a quantitative measure how particular interval is promising from the viewpoint of finding the global minimum inside this one. The inequalities (\ref{OrderedR}) arrange the intervals according to the characteristics on the ones, and the trials are performed in parallel in p most promising intervals.

\subsection{Theoretical estimates of the speedup of the parallel algorithm}
Let us describe (according to \cite{Strongin2000}) the theoretical properties of the parallel algorithm featuring its speedup. In the problems being solved, the time of executing a single search trial exceeds the time of processing its results essentially. Therefore, the speedup in the number of trials is the key characteristic of the efficiency of the parallel global search algorithm
\begin{equation} \label{par_trl_ref}
s(p) = \frac{k(1)p}{k(p)},
\end{equation}
where $k(1)$ is the number of trials executed by the sequential method and $k(p)$ is the number of trials executed by the parallel method employing $p$ parallel processes.

Obviously, the quantities of trials $n(p)$ for the algorithms with different degrees of parallelization $p$ will differ from each other. Indeed, the sequential algorithm has the full information obtained at the previous $k$ iterations when selecting the point $x^{k+1}$ of the next $(k+1)^{th}$ trial. The parallel algorithm selects not one but $p$ points $x^{k+j}$, $1 \leq j \leq p,$ on the base of the same information at the $(k+1)^{th}$ iteration. It means that the selection of the points $x^{k+j}$ is performed without the information on the trial results at the points $x^{k+i}$, $1 \leq i < j$. Only the first point $x^{k+1}$ will match to the point selected by the sequential algorithm. The points of other trials, in general, may not match the ones generated by the sequential algorithm. Therefore, we will call such trials redundant and the quantity
\begin{displaymath}
\lambda(p) = \left\{ \begin{array}{ll}
                (n(p) - n(1)) / n(p), & \textrm{$n(p) > n(1)$}\\
                0, & \textrm{$n(p) \leq n(1)$}
  \end{array} \right.
\end{displaymath}
we will call the redundancy of the method.

@@

\textbf{Theorem}. Let $x^*$ be the global minimum point and $x^{\prime}$ be the local minimum one of the function $f(x)$ and let the following conditions be satisfied:
    \begin{enumerate}
        \item The inequality is fulfilled
            \begin{equation} \label{first_s_ref}
                f(x^{\prime}) - f(x^*) \leq \delta, \delta > 0.
            \end{equation}
        \item The first $q(l)$ trials of the sequential method and of the parallel one match to each other, i.e.
            \begin{equation} \label{second_s_ref}
                \{x^1,...,x^{q(l)}\} = \{y^1,...,y^{q(l)}\}
            \end{equation}
        where
            \begin{equation} \label{third_s_ref}
                \{x^1,...,x^{q(l)}\} \subset \{x^k\}, \{y^1,...,y^{q(l)}\}\subset \{y^k\}.
            \end{equation}
        \item There exist a point $t^n \in \{y^m\}$, $n < q(l)$ such as $x^{\prime} \leq y^n \leq x^*$ or $x^* \leq y^n \leq x^{\prime}$.
        \item For the quantity $M$ from (\ref{M}), the inequality is fulfilled 
            \begin{equation} \label{fourth_s_ref}
                M > 2^{2 - 1/N} H.
            \end{equation}
        where $H$ is H{\"o}lder constant of the objective one-dimensional function.
    \end{enumerate}

Then, the parallel algorithm at $p=2$ will be non-redundant (i.e. $s(2)=2$, $\lambda(2)=0$) as long as the condition is fulfilled 
\begin{equation} \label{p_two_ref}
    (x_{t_j} - x_{t_j - 1})^{1/N} > \frac{4\delta}{M - 2^{2 - 1/N} H}, j=1,2.
\end{equation}
where $t_j$ is defined according to Step 4 of the algorithm.

\textbf{Consequence}. Let the objective function $f(x)$ have $Q$ points of local minima $\{x_1^{\prime},...,x_Q^{\prime}\}$, which the condition \eqref{first_s_ref} is fulfilled for, and let there exist the trial points $y^{n_i}$, $1 \leq i \leq Q$ such as 
\begin{gather} 
    y^{n_i} \in \{y^1,...y^{q(l)}\}, \label{conseq_range_ref} \\ 
    \alpha_i \leq y^{n_i} \leq \alpha_{i+1}, \alpha_i, \alpha_{i+1} \in \{x^*, x_1^{\prime},...,x_Q^{\prime}\}, 1 \leq i \leq Q. \label{conseq_ref}
\end{gather}
Then, if the conditions of theorem are fulfilled, the parallel algorithm with the parallelization degree $Q+1$ will be non-redundant (i.e. $s(Q+1)=Q+1, \lambda(Q+1)=0$) until the condition \eqref{p_two_ref} is fulfilled.

The consequence from Theorem plays a special role in solving the multidimen-sional problems reduced to the one-dimensional ones using the Peano-like evolvent $y(x)$. The evolvent $y(x)$, being an approximation to the Peano curve, has the effect of «splitting» of the global minimum point $y^* \in D$ into several (up to $2^N$) preimages in the interval [0,1]. Applying the parallel algorithm to minimize the reduced one-dimensional function, one can expect a zero redundancy at the parallelization degree up to $2^N+1$.

Now, let us consider the case when after the stage of the search for the rough approximation to the global solution, the stage of its local refinement is performed. Assume that the fraction of the search trials necessary for the local refinement of the solution makes $\alpha$ of total number of trials. Then, according to Amdahl's law, the overall speedup $S(p)$, which can be obtained by parallelization into $p$ processes, cannot exceed the quantity 
$$ S(p) = {\Big(\alpha + \frac{1-\alpha}{p}\Big)}^{-1}.$$

This relation will be true if one assumes the non-redundant parallelization of the global search stage. The increase in the efficiency of the parallel algorithm will be limited by the fraction of trials at the stage of the local refinement of the solution. So far, to increase the efficiency of parallelization (especially in the case of a large number of processes employed), one needs to parallelize not only the global search but the local one as well. Two particular local methods applicable to the optimization problems with the ``black-box'' type functions as well as the methods of parallelization of these ones will be considered in the next section.

\section{Local optimization methods}\label{sec:local}
\subsection{BFGS method}

BFGS (Broyden-Fletcher-Goldfarb-Shanno) method (see \cite{Nocedal}) belongs to the class of quasi-Newton methods using the values of the function $f(x^k)$ and of its gradient $\nabla f(x^k)$ to organize the search. In this method, the function f is assumed to have the properties of the quadratic model that corresponds to the existence of a symmetric matrix of the second derivatives (Hessian) of the function $\nabla^2 f(x^k)$.

The basic idea for all quasi-Newton methods consists in the construction of an approximation of the actual Hessian $\nabla^2 f(x^k)$ at every step of the algorithm using a special matrix $B^k$ followed by selection of the descent direction $d^k$ as a solution of the following system:
$$ B^k d^k = -\nabla f(x^k) \leftrightarrow d^k = -H^k \nabla f(x^k),$$
where $H^k={(B^k)}^{-1}$.

After obtaining the direction, a step is performed from current point $x^k$:
$$ x^{k+1}=x^k+ \alpha_k d^k,$$
where $\alpha_k>0$ is the adjustable step length.

Assuming $B^0=E$, the Hessian approximation is recalculated by the formula $B^{k+1}= B^k+ U^k$ where $U^k$ are the corrections represented in the form of a matrix of the rank 1. The small rank of the corrections $U^k$ is necessary for efficient procedure of calculating the inverse matrix $H^{k+1}=(B^{k+1})^{-1}=(B^k+ U^k)^{-1}$.

The main requirement to forming the corrections $U^k$ of particular form is the secant equation $B^{k+1} (x^{k+1} - x^k )= \nabla f(x^{k+1})- \nabla f(x^k)$, which is performed for all quasi-Newton methods. For more precise determination of the matrix $B^{k+1}$, it is necessary to impose additional requirements on the Hessian approximation. The set of such requirements is determined by BFGS rule. Assume:
$$ s_k = x^{k+1} - x^k,$$
$$ y_k = \nabla f(x^{k+1})- \nabla f(x^k).$$

Then, the BFGS recalculation scheme takes the following form:
$$ B^{k+1} = B^k - \frac{B^k s_k s_k^T B^k}{s_k^T B^k s_k} + \frac{y^k (y^k)^{T}}{(y^k)^{T} s_k},$$
$$ H^{k+1} = (E - \rho_k s_k (y^k)^{T}) H^k (E - \rho_k y^k s_k^T) + \rho_k s_k s_k^T,$$
where $\rho_k = 1/((y^k)^{T} s_k)$.

For the systems with limited memory, a modification of BFGS method – the limited BFGS in the limited space (L-BFGS-B) is used. The modified method stores the history of $\aleph^k = ((s_{k-i}, y^{k-i}))_i^l$ of the last $l$ vectors $s_k$ and $y^k$. From the history, the approximate form of the matrix $H^k$ is restored.

The algorithm of the BFGS method can be represented by a sequence of the fol-lowing commands:

@@

Step 6. Stop, if the condition $\|\nabla f^k\| < \varepsilon$ is fulfilled.

\subsection{Hooke-Jeeves method}
Hooke-Jeeves method (see, e.g., \cite{Himmelblau}) is the first-order one, i.e. it needs only the function values for the work. In this method, the search for the minimum at every step is performed as a result of a shift along some sample direction (step by a sample), which is constructed and then corrected as a result of special trial coordinate translations called constructing a configuration.

@@

if $\varphi(\bar{y} + he^i) < \varphi(\bar{y})$ then set $\bar{y} = \bar{y} + he^i$ else if $\varphi(\bar{y} - he^i) < \varphi(y)$ then $\bar{y} = y - he^i.$

Next, the step-by-step description of the whole method is given.

@@

The meaning of the actions performed within the steps 2, 3, and 4 can be explained as follows. 
Step 4 was introduced in order to enable the method to increase the shift magnitude within a single iteration rapidly in the case when the point $t^k$ is located far away from the solution. At that, prior to make $y^{k+1}$ current point of the iteration, a configuration is constructed from the point $y^{k+1}$. By these means, in the case of obtaining a point with $\varphi(t^{k+1})< \varphi(t^k)$, the next step by sample, in general, will be performed in a changed direction (relative to previous one) that allows adapting the method to variations in the function relief.

Finally, upon obtaining a value $\varphi(t^{k+1}) \geq \varphi(t^k)$ at Step 3, an attempt to run the algorithm again from the point $t^0=t^k$ (better than the one found) is undertaken. At that, if such an attempt hasn’t been undertaken yet, the parameter $h$ is not changed, but if it has been undertaken already, $h$ is cut in half first.

\subsection{Parallelization of the local optimization methods}

@@

For example, in Hooke-Jeeves method one can parallelize the construction of configuration i. e. parallelize the loop at the second step of the algorithm. In this loop, one may not compute the function values at the points $\bar{y} + he^i$ or $\bar{y} - he^i$ immedeatedly but may store the coordinates of the ones in the buffer, and upon accumulating $P$ points compute the function values at these ones in parallel. Afterwards, the accumulated information is analyzed and the coordinates of the point $\bar{y}$ are changed. This operation is repeated until all $N$ coordinates $\bar{y}$ are computed.

@@

\section{Results of numerical experiments}
The numerical experiments were carried out on Lobachevsky supercomputer. Each supercomputer node was a two-processor one (two Intel Sandy Bridge E5-2660 2.2 GHz, 64 Gb RAM). All methods considered in the above sections were implemented using Globalizer software \cite{globalizerSystem} (development language – C++). Intel C++ 17.0.0 compiler was used to assembly the system.
